---
title: "Intro to spatial object"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}

library(sp)
library(sf)
```


## Conceptualizing spatial vector objects in R

The foundational structure for any spatial object in {sp} is the `Spatial` class with two slots

- a bounding box
- a CRS class object to define CRS (Coordinate Reference System)

To manually build up a spatial object, the steps are:

1. Create geometric objects (Points, Lines, Polygons)

- Points: generated out of either a single coordinate or set of coordinates, like two column matrix or dataframe with columns for latitude and longitude.

- Lines: generated out of `Line` objects and `Line` object is a collection of 2D coordinates and generated out of a two column matrix or dataframe with one column for latitude and one column for longitude. so a `Lines` object is a list of one or more `Line` objects, for example all the contours at a single elevation.

- Polygons: generated out of `Polygon` objects which is a collection of 2D coordinates with equal first and last coordinates and is generated out of a two column matrix and dataframe with columns for latitude and longitude. So a `Polygons` object is a list of one or more `Polygon` objects, say, islands belonging to the same country

2. Create spatial objects (SpatialPoints, SpatialLines, SpatialPolygons)

This step adds the bounding box (automatically) and slot for CRS (which needs to be filled manually)

3. Add attributes (Optional)

If we add a dataframe with attribute data, then SpatialPoints would be a SpatialPointsDataFrame object, similarly we would have SpatialLinesDataFrame, SpatialPolygonsDataFrame objects


## Example (creating a sp object that contains highways)

```{r sp_example}

# First we would create a `Line` object that holds one highway.
ln1 <- Line(matrix(runif(6), ncol = 2)) # 1st highway
ln2 <- Line(matrix(runif(6), ncol = 2)) # 2nd highway

# then create the Lines object for the two highway
lns1 <- Lines(list(ln1), ID = c("hw1"))
lns2 <- Lines(list(ln2), ID = c("hw2"))

# turn lines into a geospatial object
sp_lns <- SpatialLines(list(lns1, lns2))

dfr <- data.frame(id = c("hw1", "hw2"),
                  cars_par_hour = c(78, 22))

sp_lns_dfr <- SpatialLinesDataFrame(sp_lns, dfr, match.ID = "id")
```


## The {sf} package

`sf` implements a formal standard called "Simple features" that specifies the storage and access model of Spatial Geometries (point, line, polygon)

In `sf` spatial objects are stored as a simple data frame with a special column that contains the information for the geometry coordinates. That special column is a list with the same length as the number of rows in the data frame and each of the individual elements of that column can by of any length needed to hold the coordinates that corresponds to an individual feature.

To create a `sf` object manually, the steps are:

1. Create geometric objects (point, line, polygon)

- Geometric object (simple features) can be created from a numeric vector, matrix or a list with the coordinates. They are called `sfg` (simple feature geometry). we can use `st_point()`, `st_linestring()`, `st_polygon()` to create these simple features.

2. Combine all the individual single feature objects for the special column

- feature geometries are combined into a simple feature collection (sfc) using `st_sfc()`. The sfc object also holds the bbox and projection information.

3. Add attributes with st_sf()

- st_sf() extends the well known R data.frame with a column that hold sfc


## Example (creating a sf object that contains highways)

```{r example-sf}

lnstr_sfg1 <- st_linestring(matrix(runif(6), ncol = 2))
lnstr_sfg2 <- st_linestring(matrix(runif(6), ncol = 2))

lnstr_sfc <- st_sfc(lnstr_sfg1, lnstr_sfg2)
lnstr_sfc

lnstr_sf <- st_sf(dfr, lnstr_sfc)
lnstr_sf
```

## Challange

Similarly to the example above generate a Point object in R. Use both, the sp and the sf “approach”.

1. Create a matrix pts of random numbers with two columns and as many rows as you like. These are your points.
2. Create a dataframe attrib_df with the same number of rows as your pts matrix and a column that holds an attribute. You can make up any attribute.
3. Use the appropriate commands and pts to create 
  - a SpatialPointsDataFrame and
  - an sf object with a gemoetry column of class sfc_POINT.
4. Try to subset your spatial object using the attribute you have added and the way you are used to from regular data frames.
5. How do you determine the bounding box of your spatial object?

```{r 1st-challange}

pts <- matrix(runif(10), ncol = 2)
attrib_df <- data.frame(popn = sample(50:90, size = nrow(pts)))

# using sp
sp_pts_dfr <- SpatialPointsDataFrame(pts, attrib_df)
sp_pts_dfr

# using sf
pts_sf <- cbind(attrib_df, data.frame(pts)) |> st_as_sf(coords = c(2, 3))
```


