---
title: "Spatial Data Manipulation in R"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


## Attribute Join in sf

Its the process of joining data in tabular format to a data in a format that holds geometries (point, line, polygon)

```{r csv_read}
library(sf)

ph_edu <- read.csv("data/PhillyEducation.csv")
names(ph_edu)

philly_sf <- st_read("data/Philly/")
names(philly_sf)
```

So merging by `GEOID10` which is same as `GEOID`

```{r merge}
philly_sf_merged <- merge(philly_sf, ph_edu, by.x = "GEOID10", by.y = "GEOID")

names(philly_sf_merged)
```

## Attribute joins in sp

```{r sp_join}

philly_sp <- rgdal::readOGR("data/Philly/", "PhillyTotalPopHHinc")

philly_sp_merged <- merge(philly_sp, ph_edu, by.x = "GEOID10", by.y = "GEOID")
names(philly_sp_merged)
```

## Topological subsetting (select polygons by Location)

we want to select all Philadelphia census tracts within a range of 2 kilometers from the city center.

The steps could be

1. Get the census tract polygons
2. Find the philadelphia city center coordinates
3. create a buffer around the city center point
4. select all the census tract polygons that intersects with the center buffer.

### Using sf package

So for first step, we have `philly_sf` as census tract polygons. And for the 2nd step, we need to create a `sf` point object with the philadelphia city center coordinates, which are, `x = 1750160, y = 467499.9`.

```{r point_sfc}

philly_center_sfc <- st_sfc(st_point(c(1750160, 467499.9)), crs = "ESRI:102003")
```
Next, we create a 2km buffer around the city center point.

```{r buffer}

philly_buf_sf <- st_buffer(philly_center_sfc, 2000)

philly_buf_sf
```


Now we can use that buffer to select all census tract polygons that intersect with that buffer and for that we can use `st_intersect` which returns a logical vector which we can use later for subsetting.

```{r subsetting}
philly_buf_intersects <- st_intersects(philly_buf_sf, philly_sf)

st_crs(philly_sf)[1]
st_crs(philly_buf_sf)[1]
```
```{r crs}
st_crs(philly_sf) <- "ESRI:102003"

philly_buf_intersects <- st_intersects(philly_buf_sf, philly_sf)

class(philly_buf_intersects)
```
> sgbp is a sparse geometry binary predicate. which is a list with integer vectors only holding the indices for each polygon that intersects.

In our case we have only one vector, because we only intersect with one buffer polygon, which we can extract using `philly_buf_intersects[[1]]` and use it for subsetting.

```{r}

philly_sel_sf <- philly_sf[philly_buf_intersects[[1]], ] #subsetting rows

plot(st_geometry(philly_sf), border = "#aaaaaa",
     main = "Census tracts that fall within 2km of city center")
plot(st_geometry(philly_sel_sf), add = TRUE, col = "red")
plot(st_geometry(philly_buf_sf), add = TRUE)
```

## Reprojecting

we can reproject a spatial object to a new CRS with

- a `sf` object with `st_transform()`
- a `raster` object with `projectRaster()`

We can retrieve the crs from a spatial object

- from an `sf` object with `st_crs()`
- from a `raster` object with `crs()`

```{r homicide_sf}
philly_homicides_sf <- st_read("data/PhillyHomicides/")

st_crs(philly_homicides_sf)
st_crs(philly_sf)
```

> Look in the ID at the very last of this long output

> USA_Contiguous_Albers_Equal_Area_Conic (aea) is a projected coordinate system with **numeric units**

we need to make sure both of `philly_sf`, `philly_homicides_sf` are in same CRS

```{r}
philly_homicides_sf_aea <- st_transform(philly_homicides_sf, st_crs(philly_sf))

st_crs(philly_homicides_sf_aea)
```
we can use `range()` from baseR and compare the coordinates before and after the reprojection and confirm that we actually have transformed them

```{r range}
range(st_coordinates(philly_homicides_sf))

range(st_coordinates(philly_homicides_sf_aea))
```
we also can compare visually

```{r compare_plot}
par(mfrow = c(1, 2))

plot(st_geometry(philly_homicides_sf), axes = TRUE, main = "before transform - laton")

plot(st_geometry(philly_homicides_sf_aea), axes = TRUE, main = "after transform - aea")
```

lets save this reprojected file

```{r save-file}
st_write(philly_homicides_sf_aea, "data/PhillyHomicides_aea", driver = "ESRI Shapefile")
```

